# Model Selection and Regularization

```{r, echo=FALSE}
# Unattach any packages that happen to already be loaded. In general this is unecessary
# but is important for the creation of the book to not have package namespaces
# fighting unexpectedly.
pkgs = names(sessionInfo()$otherPkgs)
if( length(pkgs > 0)){
  pkgs = paste('package:', pkgs, sep = "")
  for( i in 1:length(pkgs)){
    detach(pkgs[i], character.only = TRUE, force=TRUE)
  }
}
```

```{r, warning=FALSE, message=FALSE}
library(dplyr)    # data frame manipulations
library(ggplot2)  # plotting

library(caret)
library(glmnet)
```

## Stepwise selection using AIC 

Many researchers use forward or backward stepwise feature selection for both linear models or generalized linear models.  There are a number of functions in R to facilitate this, notatbly `add1`, `drop1` and `step`.

We have a data set from the `faraway` package that has some information about each of the 50 US states. We'll use this to select a number of usefule covariates for predicting the states Life Expectancy.

```{r}
library(faraway)
state.data <- state.x77 %>% data.frame() %>%
  mutate( State = rownames(.)) %>%
  mutate( HS.Grad.2 = HS.Grad^2,
          Income.2  = Income^2 )
# add a few squared terms to account for some curvature.
```


It is often necessary to compare models that are not nested. For example, I might want to compare 
$$y=\beta_{0}+\beta_{1}x+\epsilon$$
vs
$$y=\beta_{0}+\beta_{2}w+\epsilon$$

This comparison comes about naturally when doing forward model selection and we are looking for the “best” covariate to add to the model first.

Akaike introduced his criterion (which he called “An Information Criterion”) as
$$AIC=\underset{\textrm{decreases if RSS decreases}}{\underbrace{-2\,\log L\left(\hat{\boldsymbol{\beta}},\hat{\sigma}|\,\textrm{data}\,\right)}}+\underset{\textrm{increases as p increases}}{\underbrace{2p}}$$
where 
$L\left(\hat{\boldsymbol{\beta}}|\,\textrm{data}\,\right)$ is the likelihood function and $p$ is the number of elements in the $\hat{\boldsymbol{\beta}}$
vector and we regard a lower AIC value as better. Notice the $2p$
term is essentially a penalty on adding addition covariates so to lower the AIC value, a new predictor must lower the negative log likelihood more than it increases the penalty.

To convince ourselves that the first summand decreases with decreasing RSS in the standard linear model, we examine the likelihood function
$$\begin{aligned}
f\left(\boldsymbol{y}\,|\,\boldsymbol{\beta},\sigma,\boldsymbol{X}\right)	&=	\frac{1}{\left(2\pi\sigma^{2}\right)^{n/2}}\exp\left[-\frac{1}{2\sigma^{2}}\left(\boldsymbol{y}-\boldsymbol{X}\boldsymbol{\beta}\right)^{T}\left(\boldsymbol{y}-\boldsymbol{X}\boldsymbol{\beta}\right)\right] \\
	&=	L\left(\boldsymbol{\beta},\sigma\,|\,\boldsymbol{y},\boldsymbol{X}\right)
\end{aligned}$$
and we could re-write this as
$$\begin{aligned}
\log L\left(\hat{\boldsymbol{\beta}},\hat{\sigma}\,|\,\textrm{data}\right)	&=	-\log\left(\left(2\pi\hat{\sigma}^{2}\right)^{n/2}\right)-\frac{1}{2\hat{\sigma}^{2}}\left(\boldsymbol{y}-\boldsymbol{X}\hat{\boldsymbol{\beta}}\right)^{T}\left(\boldsymbol{y}-\boldsymbol{X}\hat{\boldsymbol{\beta}}\right) \\
	&=	-\frac{n}{2}\log\left(2\pi\hat{\sigma}^{2}\right)-\frac{1}{2\hat{\sigma}^{2}}\left(\boldsymbol{y}-\boldsymbol{X}\hat{\boldsymbol{\beta}}\right)^{T}\left(\boldsymbol{y}-\boldsymbol{X}\hat{\boldsymbol{\beta}}\right) \\
	&=	-\frac{1}{2}\left[n\log\left(2\pi\hat{\sigma}^{2}\right)+\frac{1}{\hat{\sigma}^{2}}\left(\boldsymbol{y}-\boldsymbol{X}\hat{\boldsymbol{\beta}}\right)^{T}\left(\boldsymbol{y}-\boldsymbol{X}\hat{\boldsymbol{\beta}}\right)\right] \\
	&=	-\frac{1}{2}\left[+n\log\left(2\pi\right)+n\log\hat{\sigma}^{2}+\frac{1}{\hat{\sigma}^{2}}RSS\right]
\end{aligned}$$
 

It isn't clear what we should do with the $n\log\left(2\pi\right)$ term in the $\log L()$ function. There are some compelling reasons to ignore it and just use the second, and there are reasons to use both terms. Unfortunately, statisticians have not settled on one convention or the other and different software packages might therefore report different values for AIC. 

As a general rule of thumb, if the difference in AIC values is less than two then the models are not significantly different, differences between 2 and 4 AIC units are marginally significant and any difference greater than 4 AIC units is highly significant.

Notice that while this allows us to compare models that are not nested, it does require that the same data are used to fit both models. Because I could start out with my data frame including both $x$ and $x^{2}$, (or more generally $x$ and $f\left(x\right)$ for some function $f()$) you can regard a transformation of a covariate as “the same data”. However, a transformation of a y-variable is not and therefore we cannot use AIC to compare a models `log(y) ~ x` versus the model `y ~ x`.

Another criterion that might be used is *Bayes Information Criterion* (BIC) which is

$$BIC=-2\,\log L\left(\hat{\boldsymbol{\beta}},\hat{\sigma}|\,\textrm{data}\,\right)+p\log n$$

and this criterion punishes large models more than AIC does (because $\log n>2$ for $n\ge8$)

The AIC value of a linear model can be found using the AIC() on a lm() object.

```{r}
m1 <- lm(Life.Exp ~ Income + Income.2 + Murder + Frost, data=state.data)
m2 <- lm(Life.Exp ~ Illiteracy + Murder + Frost, data=state.data)

AIC(m1)
AIC(m2)
```

Because the AIC value for the first model is lower, we would prefer the first model that includes both `Income` and `Income.2` compared to model 2, which was `Life.Exp ~ Illiteracy+Murder+Frost`.

### Adjusted `R-sq`
 
One of the problems with $R^{2}$ is that it makes no adjustment for how many parameters in the model. Recall that $R^{2}$ was defined as 
$$R^{2}=\frac{RSS_{S}-RSS_{C}}{RSS_{S}}=1-\frac{RSS_{C}}{RSS_{S}}$$
where the simple model was the intercept only model. We can create an $R_{adj}^{2}$ statistic that attempts to add a penalty for having too many parameters by defining
$$R_{adj}^{2}=1-\frac{RSS_{C}/\left(n-p\right)}{RSS_{S}/\left(n-1\right)}$$
With this adjusted definition, adding a variable to the model that has no predictive power will decrease $R_{adj}^{2}$.

### Example
Returning to the life expectancy data, we could start with a simple model add covariates to the model that have the lowest AIC values. R makes this easy with the function `add1()` which will take a linear model (which includes the data frame that originally defined it) and will sequentially add all of the possible terms that are not currently in the model and report the AIC values for each model.

```{r}
# Define the biggest model I wish to consider
biggest <- Life.Exp ~ Population + Income + Illiteracy + Murder + 
                      HS.Grad + Frost + Area + HS.Grad.2 + Income.2

# Define the model I wish to start with
m <- lm(Life.Exp ~ 1, data=state.data)

add1(m, scope=biggest)  # what is the best addition to make?
```

Clearly the additiona of `Murder` to the model results in the lowest AIC value, so we will add `Murder` to the model. Notice the `<none>` row corresponds to the model m which we started with and it has a `RSS=88.299`. For each model considered, R will calculate the `RSS_{C}` for the new model and will calculate the difference between the starting model and the more complicated model and display this in the Sum of Squares column.

```{r}
m <- update(m, . ~ . + Murder)  # add murder to the model
add1(m, scope=biggest)          # what should I add next?
```

There is a companion function to `add1()` that finds the best term to drop. It is conveniently named `drop1()` but here the `scope` parameter defines the smallest model to be considered.

It would be nice if all of this work was automated. Again, R makes our life easy and the function `step()` does exactly this. The set of models searched is determined by the scope argument which can be a *list* of two formulas with components upper and lower or it can be a single formula, or it can be blank. The right-hand-side of its lower component defines the smallest model to be considered and the right-hand-side of the upper component defines the largest model to be considered. If `scope` is a single formula, it specifies the upper component, and the lower model taken to be the intercept-only model. If scope is missing, the initial model is used as the upper model.

```{r}
smallest <- Life.Exp ~ 1
biggest <- Life.Exp ~ Population + Income + Illiteracy + 
                      Murder + HS.Grad + Frost + Area + HS.Grad.2 + Income.2
m <- lm(Life.Exp ~ Income, data=state.data)
step(m, scope=list(lower=smallest, upper=biggest))
```

Notice that our model selected by `step()` is not the same model we obtained when we started with the biggest model and removed things based on p-values. 

The log-likelihood is only defined up to an additive constant, and there are different conventional constants used. This is more annoying than anything because all we care about for model selection is the difference between AIC values of two models and the additive constant cancels. The only time it matters is when you have two different ways of extracting the AIC values. Recall the model we fit using the top-down approach was

```{r}
# m1 was
m1 <- lm(Life.Exp ~ Income + Murder + Frost + Income.2, data = state.data)
AIC(m1)
```

and the model selected by the stepwise algorithm was

```{r}
m3 <- lm(Life.Exp ~ Murder + Frost + HS.Grad + Population, data = state.data)
AIC(m3)
```

Because `step()` and `AIC()` are following different conventions the absolute value of the AICs are different, but the difference between the two is constant no matter which function we use.

First we calculate the difference using the AIC() function:

```{r}
AIC(m1) - AIC(m3)
```

and next we use `add1()` on both models to see what the AIC values for each.

```{r}
add1(m1, scope=biggest)
add1(m3, scope=biggest)
```


Using these results, we can calculate the difference in AIC values to be the same as we calculated before $$\begin{aligned}
-22.465--28.161	&=	-22.465+28.161 \\
	&=	5.696
	\end{aligned}$$

```{r}
smallest <- Life.Exp ~ 1
biggest  <- Life.Exp ~ Population + Income + Illiteracy + 
                       Murder + HS.Grad + Frost + Area 
m <- lm(Life.Exp ~ Income, data=state.data)
step(m, scope=list(lower=smallest, upper=biggest))
```

This same approach works for `glm` objects as well.  Unfortunately there isn't a way to make this work via the `caret` package, and so we can't do quite the same thing in general.


## Model Regularization via LASSO and Ridge Regression

For linear and generalized linear models, we might consider adding a penalty to the residual sum of squares function (RSS), which we seek to minimize. By minimizing RSS with the adding a penalty in the form of either $\lambda\sum |\beta_j|$ or $\lambda\sum \beta_j^2$ we get either LASSO or ridge regression. Via Lagrange multipliers, it is possible to show that this is equivlant to minimizing 

$$\textrm{LASSO}: \;\; \sum_{i=1}^n (y_i - X_{i,\cdot}\beta)^2 \; \textrm{ where } \sum_{i=1}^p | \beta_j | \le s$$
$$\textrm{Ridge Reg}: \;\; \sum_{i=1}^n (y_i - X_{i,\cdot}\beta)^2 \; \textrm{ where } \sum_{i=1}^p \beta_j ^2 \le s$$


Even though the name for ridge regression includes the term _regression_, these techniques work equally well in both the regression and classification situations. Unsurprisingly the methods for fitting these models in `caret` are similar, and will rely on the package `glmnet`. We will first present a regression example using both methods and then move to a classification example.

### Regression
For this example, we'll consider data from a study about prostate cancer and we are interested in predicting a prostate specific antigen that is highly elevated in cancerous tumors.

```{r}
data('prostate', package='faraway')
```


```{r, fig.height=4}
# Define how we will do our cross validation to select the tuning parameter
ctrl <- trainControl( method='repeatedcv', repeats=4, number=5,  
                      preProcOptions = c('center','scale'))  # center and scale the covariates first!
```

```{r}
# Define the grid of tuning parameters we will consider
grid <- data.frame( 
  alpha  = 0,  # 0 => Ridge Regression
  lambda = exp(seq(-5, 8, length=100)) )  # Figured out this range via trial and error

model <- train( lpsa ~ ., data=prostate, method='glmnet',
                trControl=ctrl, tuneGrid=grid,
                lambda= grid$lambda )   # Not sure why lambda isn't being passed in...

plot.glmnet(model$finalModel, xvar='lambda')
#autoplot(model$finalModel, xvar = 'lambda' )  # bug in ggfortify, doesn't change x axis
```
Each line corresponds to the $\beta_j$ coefficient for each $\lambda$ value. The number at the top is the number of non-zero coefficients for that particular $\lambda$ value.

Next we need to figure out the best value of $\lambda$ that we considered.
```{r, fig.height=3}
plot.train(model, xTrans = log, xlab='log lambda')
```

So based on this graph, we want to choose $\lambda$ to be as large a possible without increasing RMSE too much.  So $\log( \lambda ) \approx -2$ seems about right. And therefore $\lambda \approx e^{-2.35} = 0.095$.  However, we also know that this value is likely to change from run to run and we ought to consider a slightly larger value for $\lambda$.  It would be nice if this graph showed the standard error of the estimated RMSE.

```{r, fig.height=3}
num = 20    # We did 20 hold out sets (4x5 fold CV)
str( model$results ) # data.frame with model metrics and standard deviations

ggplot(model$results, aes( x=log(lambda), y=RMSE )) +
  geom_point(  ) +
  geom_line(  ) +
  geom_linerange(aes( ymin= RMSE - RMSESD/sqrt(num), ymax= RMSE+RMSESD/sqrt(num)), alpha=0.3) +
  scale_x_continuous(breaks=seq(-6,8, by=2))
```

Given this, I feel ok chosing anything from about $\log(\lambda) \in [-2, 0]$, but I would like something a bit more precise. One method suggested by Karl Breiman is to chose the largest value of $\lambda$ that is within 1 Standard Error of the best fitting $\lambda$. Given this, I think approximately $\log(\lambda)=-0.5$ which is $\lambda \approx 0.6$. 

```{r}
# Best tuning value
model$bestTune
best.index <- which.min( model$results[, 'RMSE'] )

# Best tuning value within 1 SE of optimal
# Num is the number of hold out sets considered.  4x5=20
bestTuneOneSE <- model$results %>%
  mutate( index = 1:n() ) %>%
  filter( RMSE <= min( RMSE + RMSESD/sqrt(num) ) ) %>%
  arrange(desc(lambda)) %>%
  slice(1)
bestTuneOneSE
bestTuneOneSE.index <- bestTuneOneSE$index
```

While `model$finalModel` contains the model fit using the $\lambda$ with the minimum RMSE, we might want to refit the using the "One SE" $\lambda$, which can be done via:
```{r}
# Grab the tuning parameters for either the best model or bestTuneOneSE
tuneGrid <- model$bestTune
tuneGrid <- bestTuneOneSE %>% dplyr::select(alpha, lambda) %>% as.data.frame()

# fit the model using the tuning parameter selected via CV.
model <- train( lpsa ~ ., data=prostate, method='glmnet',
                trControl=ctrl, tuneGrid=tuneGrid,
                lambda= tuneGrid$lambda )   # Not sure why lambda isn't being passed in...
```
and we could make predictions about a new observation using the usual `predict` command.
```{r}
data.to.predict <- data.frame(lcavol=3.5, lweight=3.2, age=55, lbph=-1.4, svi=1, lcp=2.2, gleason=6, pgg45=40)
predict(model, newdata=data.to.predict)
```


We can do a similar analysis via LASSO.
```{r}
# Define the grid of tuning parameters we will consider
grid <- data.frame( 
  alpha  = 1,  # 1 => LASSO
  lambda = exp(seq(-8, 2, length=100)) )  # Figured out this range via trial and error

model <- train( lpsa ~ ., data=prostate, method='glmnet',
                trControl=ctrl, tuneGrid=grid,
                lambda= grid$lambda )   # Not sure why lambda isn't being passed in...

plot.glmnet(model$finalModel, xvar='lambda')
#autoplot(model$finalModel, xvar = 'lambda' )  # bug in ggfortify, doesn't change x axis
```
Each line corresponds to the $\beta_j$ coefficient for each $\lambda$ value. The number at the top is the number of non-zero coefficients for that particular $\lambda$ value.

Next we need to figure out the best value of $\lambda$ that we considered.
```{r, fig.height=3}
plot.train(model, xTrans = log, xlab='log lambda')
```

So based on this graph, we want to choose $\lambda$ to be as large a possible without increasing RMSE too much.  So $\log( \lambda ) \approx -2$ seems about right. That corresponds to $exp(-2) = 0.135$. 

```{r}
# Best tuning value
model$bestTune
best.index <- which.min( model$results[, 'RMSE'] )

# Best tuning value within 1 SE of optimal
# Num is the number of hold out sets considered.  4x5=20
bestTuneOneSE <- model$results %>%
  mutate( index = 1:n() ) %>%
  filter( RMSE <= min( RMSE + RMSESD/sqrt(num) ) ) %>%
  arrange(desc(lambda)) %>%
  slice(1)
bestTuneOneSE
bestTuneOneSE.index <- bestTuneOneSE$index
```

While `model$finalModel` contains the model fit using the $\lambda$ with the minimum RMSE, we might want to refit the using the "One SE" $\lambda$, which can be done via:
```{r}
# Grab the tuning parameters for either the best model or bestTuneOneSE
tuneGrid <- model$bestTune
tuneGrid <- bestTuneOneSE %>% dplyr::select(alpha, lambda) %>% as.data.frame()

model <- train( lpsa ~ ., data=prostate, method='glmnet',
                trControl=ctrl, tuneGrid=tuneGrid,
                lambda= tuneGrid$lambda )   # Not sure why lambda isn't being passed in...
```
and we could make predictions about a new observation using the usual `predict` command.
```{r}
data.to.predict <- data.frame(lcavol=3.5, lweight=3.2, age=55, lbph=-1.4, svi=1, lcp=2.2, gleason=6, pgg45=40)
predict(model, newdata=data.to.predict)
```

Finally for LASSO, I'm really interested in which covariates are removed from the model.
```{r}
coef.glmnet( model$finalModel ) 
```
In this case, we see that age, lcp, and gleason have been removed from the model.



### Classification

For this example, we will consider a classification problem we encountered earlier.  Using the `Auto` dataset in the `ISLR` package, we will classify each vehicle make into either high or low efficiency and then look to see which covariates we should use in our model using LASSO.

```{r}
data('Auto', package='ISLR')
Auto <- Auto %>% 
  mutate( mpg_Grp = factor( ifelse(mpg > median(mpg),'High','Low'), 
                            levels=c('Low','High'))) %>%
  dplyr::select(-name, -mpg)
```

```{r}
# Define how we will do our cross validation to select the tuning parameter
ctrl <- trainControl( method='repeatedcv', repeats=4, number=5,  
                      preProcOptions = c('center','scale'),  # center and scale the covariates first!
                      classProbs = TRUE,    # So we generate phat values
                      summaryFunction = twoClassSummary )  # the summary information
# Define the grid of tuning parameters we will consider
grid <- data.frame( 
  alpha  = 1,  # 1 => LASSO
  lambda = exp(seq(-10, 2, length=100)) )  # Figured out this range via trial and error

model <- train( mpg_Grp ~ ., data=Auto, method='glmnet', 
                metric='ROC',
                trControl=ctrl, tuneGrid=grid,
                lambda= grid$lambda )   # Not sure why lambda isn't being passed in...

plot.glmnet(model$finalModel, xvar='lambda')
#autoplot(model$finalModel, xvar = 'lambda' )  # bug in ggfortify, doesn't change x axis
```

Next we need to figure out the best value of $\lambda$ that we considered.
```{r, fig.height=3}
plot.train(model, xTrans = log, xlab='log lambda', metric='ROC')
```

So based on this graph, we want to choose $\lambda$ to be as large a possible without reducing the AUC statisitc too much.  So $\log( \lambda ) \approx -1$ seems about right. That corresponds to $exp(-1) = 0.37$. 

```{r}
# Best tuning value
model$bestTune
best.index <- which.max( model$results[, 'ROC'] )

# Best tuning value within 1 SE of optimal
# Num is the number of hold out sets considered.  4x5=20
bestTuneOneSE <- model$results %>%
  mutate( index = 1:n() ) %>%
  filter( ROC >= max( ROC - ROC/sqrt(num) ) ) %>%
  arrange(desc(lambda)) %>%
  slice(1)
bestTuneOneSE
bestTuneOneSE.index <- bestTuneOneSE$index
```

While `model$finalModel` contains the model fit using the $\lambda$ with the minimum RMSE, we might want to refit the using the "One SE" $\lambda$, which can be done via:
```{r}
# Grab the tuning parameters for either the best model or bestTuneOneSE
tuneGrid <- model$bestTune
tuneGrid <- bestTuneOneSE %>% dplyr::select(alpha, lambda) %>% as.data.frame()

model <- train( mpg_Grp ~ ., data=Auto, method='glmnet',
                metric='ROC',
                trControl=ctrl, tuneGrid=tuneGrid,
                lambda= tuneGrid$lambda )   # Not sure why lambda isn't being passed in...
```
and we could make predictions about a new observation using the usual `predict` command.
```{r}
data.to.predict <- data.frame(cylinders=4, displacement=71, horsepower=65, weight=1770, acceleration=18, year=70, origin=3)
predict(model, newdata=data.to.predict)
```

Finally for LASSO, I'm really interested in which covariates are removed from the model.
```{r}
coef.glmnet( model$finalModel ) 
```
In this case, we see that LASSO has selected just cylinders and weight to include in the model.


## Exercises

1. ISLR #6.2 parts (a,b) - For parts (a) through (c), indicate which of i. through iv. is correct. Justify your answer. 
    a) The lasso, relative to least squares, is: 
        i. More flexible and hence will give improved prediction accuracy when its increase in bias is less than its decrease in variance. 
        ii. More flexible and hence will give improved prediction accuracy when its increase in variance is less than its decrease in bias.
        iii. Less flexible and hence will give improved prediction accuracy when its increase in bias is less than its decrease in variance. 
        iv. Less flexible and hence will give improved prediction accuracy when its increase in variance is less than its decrease in bias.
    b) Repeat (a) for ridge regression relative to least squares. 

2. ISLR #6.3 - Suppose we estimate the regression coefficients in a linear regression model by minimizing 
$$\sum_{i=1}^{n}\left(y_{i}-\beta_{0}-\sum_{j=1}^{p}\beta_{j}x_{ij}\right)^{2}\,\,\,\,\ \textrm{ subject to} \,\, \sum_{j=1}^{p}\left|\beta_{j}\right|\le s$$
for a particular value of s. For parts (a) through (e), indicate which of i. through v. is correct. Justify your answer. 
    a) As we increase s from 0, the training RSS will: 
        i. Increase initially, and then eventually start decreasing in an inverted U shape. 
        ii. Decrease initially, and then eventually start increasing in a U shape. 
        iii. Steadily increase. 
        iv. Steadily decrease. 
        v. Remain constant. 
    b) Repeat (a) for test RSS. 
    c) Repeat (a) for variance. 
    d) Repeat (a) for (squared) bias.
    e) Repeat (a) for the irreducible error.

3. ISLR #6.9 In this exercise, we will predict the number of applications received using the other variables in the `College` data set available in the ISLR. 
    a) Randomly split the data set into equally sized training and test sets using the following code:
    ```{r, message=FALSE, warning=FALSE}
    library(dplyr)
    data('College', package='ISLR')
    set.seed(2112)  
    train = College %>% sample_frac(0.5)
    test  = setdiff(College, train)
    ```
    b) Fit a linear model using least squares on the training set, and report estimated error rate from fitting the data to the training set as well as the error obtained from predicting values in the test set. 
    c) Fit a ridge regression model on the training set, with $\lambda$ chosen using 4x5 repeated cross-validation and select the best $\lambda$. Report the estimated error rate from the cross-validation as well as the observed test error obtained. 
    d) Fit a ridge regression model on the training set, with $\lambda$ chosen using 4x5 repeated cross-validation and select the "OneSE" $\lambda$. Report the estimated error rate from the cross-validation as well as the observed test error obtained.  
    e) Fit a lasso model on the training set, with $\lambda$ chosen using 4x5 repeated cross-validation and select the best $\lambda$. Report the estimated error rate from the cross-validation as well as the observed test error obtained.  Also report the number of non-zero coefficient estimates. 
    f) Fit a lasso model on the training set, with $\lambda$ chosen using 4x5 repeated cross-validation and select the "OneSE" $\lambda$. Report the estimated error rate from the cross-validation as well as the observed test error obtained.  Also report the number of non-zero coefficient estimates.  
    g) Comment on the results obtained. How accurately can we predict the number of college applications received? Is there much difference among the test errors resulting from these five approaches? How well does the estimated error rates correspond to the error rates you obtained when you fit your model to the test set? 

